Describe and Explore: Using feature-based time series clustering to identify meaningful structures in intensive longitudinal data

### 2022-12-18 manuscript.tex
:bug:

% Title and header
\title{Describe and Explore: Using feature-based time series clustering to understand modern intensive longitudinal data}
\shorttitle{feature-based time series clustering}

@@

% Abstract
\abstract{
Psychological time series data has not only become more common but also more diverse and complex. Researchers and practitioners are more and more interested in non-stationary developmental patterns of multivariate concepts, across different time scales, while contextualized measurements bring about issues of non-imputable missingness. At the same time, clustering analyses offer the potential of reducing these complexities to important and meaningful patterns. Such procedures could aid researchers and practitioners in describing and exploring patterns across participants and could promote the creation of more embedded theories and interventions. In this manuscript, we propose feature-based time series clustering as a flexible, transparent, and well-grounded clustering approach that addresses the growing dimensionality, missingness, and time scale challenges. We introduce the individual feature extraction, -reduction, and -clustering steps and illustrate their utility with an empirical example. We show that time series features such as means, autocorrelations, and linear trends are not only familiar to psychological researchers but are also conceptually meaningful in interpreting time series clusters. Beyond the conceptual and methodological introduction, we also provide practical algorithm overviews and readily available code for data preparation, analysis, and interpretation.

@@

\begin{mdframed}[style=mdfbox]
\noindent\center\textit{Structure}:
\begin{itemize}[nosep]
    \item introduction
    \begin{itemize}[nosep]
        \item\ [Relevance]: increase in number and variety of ESM data
        \item\ [Problem]: no descriptive / understand structure analyses
        \item\ [Proposal]: feature based clusterting = flexible +  transparent
    \end{itemize}
    \item Modern ESM data: new type of data and issues for analysis:
    \begin{itemize}[nosep]
        \item incomplete data (complete data models vs. structural missingness)
        \item non-equidistant
        \item non-stationary
        \item changes/trends on different time scales
        \item interest in multivariate developments
        \item dimensionality curse
    \end{itemize}
    \item Time series clustering
    \begin{itemize}[nosep]
        \item shape-based
        \item model-based
        \item \textbf{feature-based}
    \end{itemize}
    \item Feature-Based Clustering
    \begin{itemize}[nosep]
        \item input variables
        \item feature extraction
        \item feature reduction
        \item feature clustering
    \end{itemize}
    \item Empirical Illustration
    \begin{itemize}[nosep]
        \item Data
        \item Analysis and Results
    \end{itemize}
    \item Discussion (summary, limitation, implications, conclusion, etc.)
\end{itemize}
\vspace{1em}
\end{mdframed}

% Relevance Pragraph and issue
Recent years have seen a striking increase in the number and variety of research studies using experience sampling data \citep[][; also see \fgrref{fig:ScopusEsm}]{hamaker2017}. 
With the increased availability of technologies to easily collect large amounts of experience sampling data using mobile devices \citep[e.g.,][]{Keil2020} and web-based applications \citep[e.g.,][]{Arslan2020}, the experience-sampling method (ESM) has been applied to a host of new types of psychological data. 
This move toward a broader application of the experience-sampling method, however, stands in stark contrast with recent developments within the psychometric literature on ESM analyses, which have become more specialized and model-focused. 
While generally speaking analytical methods for ESM data have become more readily available \citep[e.g.,][]{ODonnell2021}, flexible descriptive and exploratory analyses have largely been neglected and remain crucially understudied within the experience sampling literature.

% Problem Illustration Paragraph
Importantly, the understudied `describe and explore' analyses are not only important for contextualizing inferential model tests, but are powerful methods for theoretical and applied users in their own regard. Extensive longitudinal data sets come with new forms of heterogeneity, where researchers have to consider differences across large numbers of participants, time points, and variables. Data-driven models not only describe and summarize this complex data but can help us understand the data by uncovering patterns that might not be detectable otherwise. As an example, understanding how people differ in their trajectories can be crucial in understanding adaptive and maladaptive patterns. By identifying and contextualizing distinct trajectories, we might for example determine which situation- or person-specific differences discern positive from negative developments. Outcomes from such an approach might inform both practical early-warning signals as well as academic theory-building efforts. There is, thus, a clear need to assess the utility of data-driven analysis procedures that are flexible enough to address the growing variety of intensive longitudinal psychological data.

% Aim / Solution / Proposal Paragraph:
In this article, we propose to look at recent developments within the wider time-series clustering literature to identify an exploratory data analysis that is flexible enough to apply to the growing variety of ESM data. In particular, we propose that feature-based clustering is ideally suited because it offers a flexible variety of validated and intuitive components while also allowing for adequate control by the researcher. To showcase the utility of this technique, we apply feature-based clustering to a set of three ESM studies with complex conceptualizations and missingness patterns.

\section{Modern ESM Data}
Two important developments within the psychological sciences have been a shift towards more comprehensive assessments of psychological phenomena and a growing contextualization of intensive longitudinal real-life data. Multivariate experience-sampling approaches allow researchers to test theories more exhaustively and help build a more embedded understanding of human experience processes. However, such data also deviates from past intensive longitudinal data sets that have focused on psychological states that are more homogeneous in their developmental characteristics. It, thus, makes sense to briefly consider which challenges modern ESM data introduce and what qualities a present-day clustering analysis should possess. We particularly highlight issues of dimensionality, missingness, and time scales within contemporary ESM data.

Concerning dimensionality issues, especially more abstract psychological experiences often need a wider variety of measurements to be captured adequately. Today, few clinical conditions are captured with a single symptom measure, emotions are rarely assessed in isolation, and socio-cultural experiences are now widely considered to be multimodal. This also means that modern analysis techniques increasingly need be able to accommodate an increased focus on multivariate developments. The recent rise in psychological network models has highlighted this need for more multivariate outcome measurements \citep[e.g.,][]{decancq2013}. Beyond the capability to assess multivariate developments, an increased number of variables also heightens the computational load for model estimations and, especially when the number of variables and measurements compounds, models may not converge --- an issue sometimes referred to as the dimensionality curse. A modern descriptive and exploratory analysis technique should consequently be able to summarize and structure multivariate phenomena without running into computational load issues.

The push for more contextualized data also presents two key issues with missing data. More event-based assessments tend to (1) encounter issues when models require time intervals between measurements to be identical (i.e., equidistant measurements) and an increased focus on context-specific phenomena has (2) lead to issues when data is only available under certain conditions, including event-based questions (i.e., structurally missing data). Smaller issues of non-equidistant data can be avoided with transformations \citep[e.g., dynamic time warping,][]{berndt1994} or newer modeling procedures \citep[e.g., continuous-time models][]{dehaan-rietdijk2017} but for many analyses, including many cluster approaches, non-equidistant measurements remain a prevalent issue. Structural missingness remains an even more strenuous challenge. Structural missingness occurs when data is missing because it logically cannot be collected. It would not be sensible to measure the quality of an event for participants that did not experience the event. This type of missingness particular becomes an issue when researchers are interested in how two concepts co-develop and one of the concepts has structural missingness under certain conditions. The most common practice for structurally missing data is to either exclude the variable or any measurement that has no structurally missing data \citep[e.g.,][]{lavori2008}\footnote{This is the case because the most commonly used models require complete data \citep{schafer2002} and structurally missing data cannot be imputed as it logically does not exist \citep[e.g.,][]{lavori2008}.} --- neither option suits a research question that includes a developmental relationship between the two variables. In short, adequate descriptive approaches should be able to deal with non-equidistant and structurally missing data in order to address modern ESM data.

Finally, new approaches to ESM data have highlighted two main issues with how temporal developments are conceptualized. Researchers are (1) increasingly interested in how mean-levels of variables change over time, often in a non-linear fashion (i.e., non-stationary), and (2) there is increased interest in relationships between variables that change on fundamentally different time scales. Many of the commonly used analyses for ESM data assume that means and variances are stable over time\footnote{These models include the popular vector autoregressive models, dynamic structural equation models, autoregressive integrated moving average models, and cross-lagged panel analyses. Notable time-varying adaption include \citet{haslbeck2021a}.}. While several trend modeling methodologies exist (e.g., spline regression models, latent growth curve modeling), their availability is limited and they seldomly extend into the descriptive realm. Moreover, among the approaches that allow for means to change, even fewer allow for substantially different time scales between variables, which can be essential in understanding how daily experiences relate to slower psychological phenomena that develop over days, weeks, and months \citep[e.g., see][]{Ram2014}. A modern descriptive and exploratory methodology, including a clustering approach, should thus be able to flexibly capture and compare non-stationary and non-linear trends, also across different time scales.

\section{Time Series Clustering}
One promising approach to describing and exploring time-series data has been the idea of structuring data using clustering analyses. Clustering seeks to organize data into groups that are maximally different from one another while minimizing differences within the groups. Time-series clustering has been extremely common in other empirical fields, including analyses of astronomical, meteorological, and aviation pathways, biological and medical developments, as well as energy and finance patterns \citep{Aghabozorgi2015}. But also for psychological data, time-series clustering is not entirely novel \citep[e.g.,][]{ernst2021}. However, most clustering introduced for psychological data has faced the dimensionality-, missingness-, and time scale issues described above.

Many of the common time-series clustering algorithms used today have been developed within the machine learning literature and have commonly been separated into three different approaches: (1) shape-based, (2) model-based, and (3) feature-based clustering \citep{hautamaki2008, liao2005}. Shape-based clustering analyses use the raw time series and seek to match time series by non-linearly stretching and contracting the time axes. The other two approaches do not use the raw time series directly but convert the data first. The model-based clustering has been the most commonly used method within the psychological literature and clusters the parameters of a parametric model fitted for each participant \citep[e.g.,][]{ernst2021}. The flexibility of such clustering is restricted to the assumptions of the model used, so that stationary models, for example, do not capture non-stationary data well. Finally, feature-based clustering analyses are the most flexible, as they allow users to summarize the time series data with a wide variety of features that capture the data characteristics of interest. The participants are then clustered on basis of the extracted features. As an illustration of the flexibility of this approach, a commonly used software package for this approach allows users to extract a total of 794 time series features \citep[][]{christ2018}. Time series features can be chosen by the user based on their research question and can, for example, include the mean and standard deviation of a time series, but also linear and non-linear trends, inertia-related features such as auto-correlations, or the stability of the time series (e.g., mean absolute change). Given the flexibility of the feature-based approach, users are able to apply the analysis across multivariate data, feature extraction mitigates issues of non-equidistant and structurally missing data, and the features can capture and compare across non-linear trends at different time scales.

@@

The feature-based clustering can be structured into four main steps. (1) The selection and preparation of the input variables, (2) the extraction of the features that describe the time series, (3) an optional feature reduction step if there are too many data points for the clustering algorithms, and (4) the actual clustering of the time series features. We provide a conceptual overview that can be used alongside this section in \fgrref{fig:TSCFlow}.

@@

Fortunately, past conceptual and empirical efforts offer valuable discussions of common time series features in psychological research. While the final selection of features should always be driven by the research questions and field-specific conventions, we can build a practical toolbox of meaningful time series features for psychological data. In particular, we propose to focus on six features based on common psychological research questions and recent works on affect dynamics \citep[e.g.,][]{dejonckheere2019, kuppens2017}. An overview of the proposed features, their substantive interpretations, and mathematical operationalizations is available in \tblref{tab:esmFeatures}.

@@

Each of these approaches can be a valuable clustering approach for time series feature data and users will usually have to make an informed decision based on the structure of their data as well as an appropriate weighing of accuracy and efficiency. As an example, under ideal conditions, all approaches are likely to suggest very similar clusters. However, when the shapes of clusters, for example, become more complex in a multi-dimensional space, density-based or hierarchy-based approaches that allow for more bottom-up clustering are likely to be more accurate. Yet, with higher numbers of participants and features, both density- and hierarchy-based approaches may perform less well and the more efficient centroid-based methods might be more effective. Similarly, several of the proposed time series features are statics that are generated from Gaussian distributions and a distribution-based algorithm might be ideally suited to separate such distributions. Yet, in cases where mis-specifications are costly, a method with fewer assumptions might be more advisable.

@@

@@

\section{Placeholder Title}
\label{app:AppendixTitle}
To be written, if necessary.

>>>>

% Title and header
\title{Describe and Explore: Using feature-based time series clustering to identify meaningful structures in intensive longitudinal data}
\shorttitle{feature-based time series clustering}

@@

% Abstract
\abstract{
Psychological time series data has not only become more common but also more diverse and complex. Researchers and practitioners are increasingly interested in non-stationary developmental patterns of multivariate concepts, across different time scales, while contextualized measurements bring about issues of non-imputable missingness. At the same time, clustering analyses offer the potential of reducing these complexities to important and meaningful patterns. Such procedures aid researchers and practitioners in describing and exploring patterns across participants and can promote the creation of more embedded theories and interventions. In this manuscript, we propose feature-based time series clustering as a flexible, transparent, and well-grounded clustering approach that addresses the growing challenges of dimensionality, missingness, and time scales. We introduce the individual feature extraction, -reduction, and -clustering steps and illustrate their utility with an empirical example. We show that time series features such as means, autocorrelations, and linear trends are not only familiar to psychological researchers but are also conceptually meaningful in interpreting time series clusters. Beyond the conceptual and methodological introduction, we also provide practical algorithm overviews and readily available code for data preparation, analysis, and interpretation.

@@

% \begin{mdframed}[style=mdfbox]
% \noindent\center\textit{Structure}:
% \begin{itemize}[nosep]
%     \item introduction
%     \begin{itemize}[nosep]
%         \item\ [Relevance]: increase in number and variety of ESM data
%         \item\ [Problem]: no descriptive / understand structure analyses
%         \item\ [Proposal]: feature based clusterting = flexible +  transparent
%     \end{itemize}
%     \item Modern ESM data: new type of data and issues for analysis:
%     \begin{itemize}[nosep]
%         \item incomplete data (complete data models vs. structural missingness)
%         \item non-equidistant
%         \item non-stationary
%         \item changes/trends on different time scales
%         \item interest in multivariate developments
%         \item dimensionality curse
%     \end{itemize}
%     \item Time series clustering
%     \begin{itemize}[nosep]
%         \item shape-based
%         \item model-based
%         \item \textbf{feature-based}
%     \end{itemize}
%     \item Feature-Based Clustering
%     \begin{itemize}[nosep]
%         \item input variables
%         \item feature extraction
%         \item feature reduction
%         \item feature clustering
%     \end{itemize}
%     \item Empirical Illustration
%     \begin{itemize}[nosep]
%         \item Data
%         \item Analysis and Results
%     \end{itemize}
%     \item Discussion (summary, limitation, implications, conclusion, etc.)
% \end{itemize}
% \vspace{1em}
% \end{mdframed}

Recent years have seen a striking increase in the number and variety of research studies that follow participants' everyday experiences and collect real-world psychological time series \citep[e.g., experience sampling method (ESM),][; also see \fgrref{fig:ScopusEsm}]{hamaker2017}. These extensive longitudinal data sets come with new forms of heterogeneity, where researchers have to consider differences across large numbers of participants, time points, and variables \citep[e.g.,][]{cattell1966, Monden2015, wardenaar2013}. Oftentimes, however, researchers are interested in precisely this complexity and wish to understand how people differ in their temporal developments across a number of distinct variables \citep[i.e., understand the structure of between-subject differences of within-participant dynamics; e.g.,][]{ernst2021}. One relatively recent addition to the toolbox of ESM researchers has been time series clustering --- the idea of grouping participants based on the characteristics of their time series dynamics \citep{ariens2020}. In practice, this means that participants are inductively grouped in such a way that the dynamics of some variable(s) have maximum similarity for individuals in the same group, and minimum similarity for individuals in different groups \citep[e.g.,][]{liao2005}.

Time series clustering has a number of conceptual use cases. Specifically, time series clustering aids in reducing the time, variable, and person complexity by extracting structure between participants. These reduction and structuring qualities can be essential in detecting phenomena and extracting more abstract functional principles \citep[][]{eronen2021a}. Such meaningful psychological principles and phenomena have often been recognized as being imperative for building robust theories and real-world interventions \citep[e.g.,][]{borsboom2021, eronen2020}.

%Time-series clustering has been extremely common in other empirical fields, including analyses of astronomical, meteorological, and aviation pathways, biological and medical developments, as well as energy and finance patterns \citep{Aghabozorgi2015}. But also for psychological data, time-series clustering has recently become more common \citep[e.g.,][]{ernst2021}.

Thus far, one of the most common clustering approaches for ESM data has been to group individuals based on person-specific model parameters --- notably intercepts and slopes from vector autoregression models \citep[VAR; e.g.,][]{ariens2020}. While these model parameters are familiar to researchers in the field, the model-based approach has a number of important shortcomings. Computationally, the number of parameters compounds with increases in variables and lag orders, which tends to result in efficiency and accuracy reduction for most clustering algorithms and is sometimes referred to as the dimensionality curse \citep{altman2018}\footnote{As an illustration of this scalability issue, consider a VAR model of 8 variables that includes the first three lags (i.e., VAR(3)). For each of the 8 variables, we would have 25 parameters --- one intercept and 24 slopes (8 variables * 3 lags each). That compounds to a total of 200 parameters per person (8 variables * 25 parameters).}. But more importantly, using model parameters means that the clustering is restricted by the assumptions and limitations of the statistical model. To take the common VAR model as an example, the model assumes that the time intervals between measurements are identical (i.e., equidistant measurements), it does not allow for missing values, and it assumes that the means as well as variances of the variables do not change over time \citep[i.e., stationarity assumption;][]{lutkepohl2005}.

These shortcomings, however, stand in sharp contrast with the types of data researchers commonly collect and common research interests in the field. Psychological researchers might, for example, collect data based on the occurrence of natural events, which tends to result in non-equidistant measurements \citep[e.g.,][]{myin-germeys2018, hamaker2017} and context-specific missingness that cannot be imputed, for instance, interaction quality perceptions \citep[e.g.,][]{kivela2022, lavori2008}. At the same time, researchers are often specifically interested in non-stationary developments, looking, for example, at how symptom levels and -variations change over time --- often abruptly and non-linearly \citep[][]{bringmann2018b, helmich2020a}. We offer a more extensive discussion of the current challenges that clustering approaches should address in \appref{app:ChallengesAppendix}. In short, the time series clustering repertoire should be extended to approaches that can more flexibly allow for diverse measurement and missingness patterns and can capture non-stationary data, including non-linear trends, also across different time scales.

Importantly, model-based clustering is not the only clustering approach, and model intercepts and slopes are not the only way in which researchers conceptualize and describe time series. A flexible extension of model-based clustering is feature-based clustering \citep[e.g.,][]{Aghabozorgi2015}\footnote{It should be noted that a third option is shape-based clustering, which uses the raw time series and seeks to match different time series by non-linearly stretching and contracting the time axes. Shape-based clustering might be less relevant to many psychological time series because the forced alignment and scaling can, for example, lead to misspecifications \citep[e.g., ][]{faloutsos1994}. Such scaling issues are less of a concern for highly regular and comparable time series such as EEG \citep[][]{huang1985}.}. Whereas model-based clustering is specifically limited to model parameters as time series characteristics, feature-based clustering extends the time series characteristics to virtually any numerical summary of a time series \citep[e.g.,][]{liao2005}. As an illustration of the flexibility of this approach, a commonly used software package for this approach allows users to extract a total of 794 time series features \citep[][]{christ2018}. Time series features can be chosen by the user based on their research question and can, for example, include the mean and standard deviation of a time series, but also linear and non-linear trends, inertia-related features such as auto-correlations, or the stability of the time series \citep[e.g., mean absolute change; e.g.,][]{barandas2020}. Given the flexibility of the feature-based approach, users are able to apply the analysis across multivariate data, feature extraction mitigates issues of non-equidistant and structurally missing data, and the features can capture and compare across non-linear trends at different time scales.

\section{Features of Psychological Time Series}

@@

------------------------

NOTES:

model based clustering: \citep{bulteel2016, stefanovic2022}

sort of feature-based clustering in psych \citep{heylen2016}

clustering in reduced space \citep{timmerman2013}

------------------------

The feature-based clustering can be structured into four main steps \citep{rasanen2009, wang2006}. (1) The selection and preparation of the input variables, (2) the extraction of the features that describe the time series, (3) an optional feature reduction step if there are too many data points for the clustering algorithms, and (4) the actual clustering of the time series features. We provide a conceptual overview that can be used alongside this section in \fgrref{fig:TSCFlow}.

@@

Fortunately, past conceptual and empirical efforts offer valuable discussions of common time series features in psychological research. While the final selection of features should always be driven by the research questions and field-specific conventions, we can build a practical toolbox of meaningful time series features for psychological data. In particular, we propose to focus on six features based on common psychological research questions and recent works on affect dynamics \citep[e.g.,][]{dejonckheere2019, kuppens2017, adya2001}. An overview of the proposed features, their substantive interpretations, and mathematical operationalizations is available in \tblref{tab:esmFeatures}.

@@

Each of these approaches can be a valuable clustering approach for time series feature data and users will usually have to make an informed decision based on the structure of their data as well as an appropriate weighing of accuracy and efficiency. As an example, under ideal conditions, all approaches are likely to suggest very similar clusters. However, when the shapes of clusters, for example, become more complex in a multi-dimensional space, density-based or hierarchy-based approaches that allow for more bottom-up clustering are likely to be more accurate. Yet, with higher numbers of participants and features, both density- and hierarchy-based approaches may perform less well and the more efficient centroid-based methods might be more effective. Similarly, several of the proposed time series features are statics that are generated from Gaussian distributions and a distribution-based algorithm might be ideally suited to separate such distributions \hl{(Corduas \& Piccollo, 2008)}. Yet, in cases where mis-specifications are costly, a method with fewer assumptions might be more advisable.

@@

\subsection{Cluster Performance Evaluation}

PARAGRAPH ON PERFORMANCE METRICS (incl., cross-validation?)

PARAGRAPH ON REPORTING \citep[e.g.,][]{vandeschoot2017}

@@

\section{ESM Data Challenges}
\label{app:ChallengesAppendix}
We will briefly consider which challenges modern ESM data introduce and what qualities are called for in an extension of the clustering repertoire. We particularly highlight issues of dimensionality, non-equidistant or missing measurements, an interest in non-stationary trends, as well as inconsistent/diverse time scales.

Concerning dimensionality issues, especially more abstract psychological experiences often need a wider variety of measurements to be captured adequately. Today, few clinical conditions are captured with a single symptom measure \citep[e.g.,][]{cramer2016}, emotions are rarely assessed in isolation \citep[e.g.,][]{reitsema2022}, and socio-cultural experiences are now widely considered to be multimodal \citep[e.g.,][]{Kreienkamp2022d}. This also means that modern analysis techniques increasingly need be able to accommodate an increased focus on multivariate developments. At the same time, however, an increase in the number of considered variables tends to come at the expense of computational load for model estimations and clustering models may not converge \citep[the aforementioned dimensionality curse;][]{altman2018}. A modern time series clustering technique should consequently be able to summarize and structure multivariate phenomena without running into computational load issues.

Another common type of data that are measurement regiments that collect data in irregular time intervals (i.e., non-equidistant measurements). Common are, for example, procedures where participants are asked to respond at random times throughout the day (i.e., signal-contingent) or following specific natural events of interest \citep[i.e., event-contingent; see][]{shiffman2008, myin-germeys2018}. Under such conditions data tends to violate the equidistance assumption that is expected by many time series models \citep[][]{hamaker2017}. Smaller issues of non-equidistant data can be avoided with transformations \citep[e.g., dynamic time warping,][]{berndt1994} or newer modeling procedures \citep[e.g., continuous-time models;][]{dehaan-rietdijk2017} but for many analyses, including some cluster approaches, non-equidistant measurements remain a prevalent issue.

Structural missingness remains an even more strenuous challenge. Structural missingness occurs when data is missing because it logically cannot be collected \citep[as opposed to probabilistically missing data;][]{little2020, mclean2017}. Often, however, researchers might want to include variables in their models that are not be available under all conditions. Follow-up and event-contingent questions are a common example in ESM studies. Researchers, for example, ask about the frequency, intensity, or duration of symptoms --- but only if a symptom was present \citep{kivela2022}. Such approaches become specifically critical in cases of sensitive questions such as questions about suicidal ideation or other potentially trauma inducing questions \citep[e.g.,][]{glenn2022}. The most common practice for structurally missing data is to either exclude the variable or any measurement that has no structurally missing data \citep[e.g.,][]{lavori2008}\footnote{This is the case because the most commonly used models require complete data \citep{schafer2002} and structurally missing data cannot be imputed as it logically does not exist \citep[e.g.,][]{lavori2008}.} --- neither option suits a research question that wishes to include variables with common structural missingness, such as event-specific or follow-up questions. In short, new clustering approaches should be able to deal with structurally missing data in order to address modern ESM data.

When it comes to studying developmental trajectories, psychological researchers are often also interested in nonstationary processes because they are more representative of the complex, dynamic behavior of the human mind. In psychology, nonstationary processes are typically used to study phenomena such as cognitive development \citep[][]{quartz1997}, decision-making \citep[][]{ratcliff2016}, and emotion dynamics \citep[][]{bringmann2018b}. These processes are often characterized by changes in the underlying statistical properties of the data over time, such as changes in the mean or variance \citep[][]{molenaar2009}. Especially when considering changes in mean levels, researchers are often interested in nonlinear changes because they describe human functioning better. For example, in decision making people might switch between choices \citep[][]{ratcliff2016}, or patients reducing medication might experience mood swings \citep[][]{helmich2020a}. Similarly, psychologists are often also interested in how variances change over time. This is especially the case because several changes in an individual's variance have been found to be indicative of critical changes, including depression relapses and symptom shifts more generally \citep[e.g.,][]{schreuder2020, wichers2020}. There is, thus, also a need for time series clustering algorithms that capture nonstationary processes, including nonlinear trends.

Psychological time series often exhibit complex patterns and relationships that can change over different time scales. For example, a time series of daily mood ratings may show a weekly pattern, with higher ratings on the weekends and lower ratings during the week. At the same time, the series may also exhibit a longer-term trend, with overall mood levels increasing or decreasing over the course of several months or years \citep[e.g.,][]{Ram2014}. \sout{These different time scales can be studied separately or in combination, using different statistical techniques and modeling approaches \citep[][]{bertenthal2007, jeronimus2019a}.} Different time scales can become an even more difficult issue when different variables in a model develop on different time scales \citep{bringmann2022b}. Different time scales are thus also a concern clustering approaches should be able to address.
